\section{Preliminaries}\label{sec:preliminaries}

This section provides the theoretical background of the proposed method. We start with the definition of Dynamic Mode Decomposition (DMD) and its online and extended online versions. We then describe how to utilize the online Singular Value Decomposition (SVD) algorithm, which enables finding lower rank representation less expensively. Finally, we present the proposed method for truncating the DMD matrix to a lower rank online.

\subsection{Dynamic Mode Decomposition: Background}
A comprehensive review of Dynamic Mode Decomposition (DMD)~\cite{Schmid2010}, DMD with control (DMDc)~\cite{Proctor2016}, its online variant (oDMDc)~\cite{Zhang2019}, is provided in Appendix~\ref{app:dmd}. DMD forms the mathematical foundation for our approach, enabling the extraction of interpretable, low-dimensional models from high-dimensional time-series data. In this work, we build upon these established techniques to develop a novel, adaptive, and interpretable change-point detection framework.

\subsection{Truncating Online DMD with Control}\label{sec:truncating-online-dmd}
Some of the challenges of oDMDc proposed in~\cite{Zhang2019} include the lack of robustness to noise, bad scalability, and decreased numerical stability of small eigenvalue updates. To address these issues, we propose modifying the oDMDc algorithm that truncates the DMD matrix to a lower rank. Conventionally, this process in batch-trained DMD relies on the truncated singular value decomposition (SVD) method, which is widely used in data analysis to reduce the dimensionality of data while preserving the most essential information. Nevertheless, computing the SVD of the matrix \(X_k\) is computationally expensive and unsuitable for online learning. Instead, we employ online SVD algorithms that perform low-rank updates of the SVD as new snapshots \(X_k\) become available.

We use the algorithm of~\cite{Zhang2022}, a modified version of the originally proposed algorithm in~\cite{Brand2006}. The main benefit of this modification is the reorthogonalization rule, which prevents erosion of left singular values orthogonality at a reasonable computational cost. For the details on the algorithm, please refer to the original work of author~\cite{Zhang2022}. For consistency of nomenclature, we will refer to the SVD decomposition of the augmented snapshot matrix \(\bar{X}_k\) as
\begin{equation*}
    \bar{X}_k = \tilde{U}_k \tilde{\Sigma}_k \tilde{V}_k^\top.
\end{equation*}

The new snapshots \(\bar{X}_{k+c}\) may be used for updating the Online SVD as shown in Algorithm~\ref{alg:online-svd-updates} in Appendix~\ref{app:dmd}. Old snapshots may be reverted using Algorithm~\ref{alg:online-svd-reverts} in Appendix~\ref{app:dmd}.

We propose incorporating the truncation using online SVD into the online DMD algorithm described in Appendix~\ref{app:online-dmd}. The truncation of the DMD matrix requires a data transformation step before updating the DMD matrix. This transformation is performed by projecting the snapshots onto the first \(r\) POD modes as
\begin{equation}
    \{\tilde{\bar{X}}_k, \tilde{X}^\prime_k\} = \{\tilde{U}^\top_k\bar{X}_k, \tilde{U}^\top_k(:m, 1:p)X^\prime_k\}.
\end{equation}

We wish to update reduced-order matrix \(\tilde{\bar{A}}_k\), which corresponds to augumented state-input matrix \(\bar{A}_k\) in~\eqref{eq:augmented-matrix}. \(\tilde{\bar{A}}_k\) is a rectangular matrix of size \(p \times r\), where \(r = p + q\), \(p\) is the rank of the reduced-order state matrix \(\tilde{A}_k\) and \(q\) is the rank of the reduced-order control matrix \(\tilde{B}_k\).

Assuming we updated online SVD on snapshots \(\bar{X}_k\), we wish to inform reduced-order matrices \(\tilde{\bar{A}}_k\) and \(\tilde{P}_k\) defined in~\eqref{eq:aux-matrices} about the change of rotation in scaled coordinate space (column space; the orthonormal basis of features). The change of rotation as new data becomes available can be tracked as \(\uis{K}{U'U}{k} = \tilde{U}_k^\top \tilde{U}_{k-1}\).

To align reduced-order matrices \(\tilde{\bar{A}}_k\) and \(\tilde{P}_k\) with this change in column space, first we decouple \(\uis{K}{U'U}{k}\) as follows:
\begin{equation}
    \uis{K}{U'U}{k} = \begin{bmatrix}
        \uis{K}{U'U}{k, p \times p} & \uis{K}{U'U}{k, p \times q} \\
        \uis{K}{U'U}{k, q \times p} & \uis{K}{U'U}{k, q \times q}
    \end{bmatrix},
\end{equation}

and then apply alignment to the reduced-order matrices as

\begin{align}
    \tilde{\bar{A}}_k & = \begin{bmatrix} \uis{K}{U'U}{k, p \times p} \tilde{A}_k \uis{K}{U'U}{k, p \times p}^\top & \uis{K}{U'U}{k, p \times p} \tilde{B}_k \uis{K}{U'U}{k, q \times q} \end{bmatrix} , \\
    \tilde{P}_k       & = {(\uis{K}{U'U}{k} P_k^{-1} \uis{K}{U'U}{k}^\top)}^{-1}.
\end{align}


What follows, is the update of reduced matrices \(\tilde{A}_k\) and \(\tilde{P}_k\) using truncated snapshots \(\tilde{X}_k\) and \(\tilde{X}^\prime_k\). The updates could be performed conveniently using formulae in~\eqref{eq:precision-matrix-update} and~\eqref{eq:online-dmd-update}.

While the computational cost of a mini-batch update is the same as applying the rank-one updates \(c\) times in the original formulation in Appendix~\ref{app:online-dmd}, the mini-batch updates of the proposed truncated online DMD yield benefits in computational cost.
