Modern control systems in industry, energy, and infrastructure increasingly rely on real-time monitoring to ensure safety, reliability, and efficiency. Detecting abrupt or gradual changes in system behavior—known as change-point detection (CPD)—is critical for preventing failures and optimizing performance, yet remains challenging in nonlinear, high-dimensional, and time-varying environments. Existing CPD methods often assume i.i.d.,  time-invariant dynamics, or lack transparency, limiting their deployment in safety-critical applications.

Here we introduce ODMD-CPD, an Adaptive and Interpretable framework based on Dynamic Mode Decomposition for change-point detection. ODMD-CPD integrates a self-supervised online learning mechanism with truncated online dynamic mode decomposition with control (toDMDc), enabling real-time adaptation to evolving system dynamics without the need for labeled change points. The method produces linear representation of the system dynamics, interpretable detection statistics and feature attributions, supporting human decision-making.

% We show that ODMD-CPD achieves provable convergence, bounded false alarm rates, and guaranteed detection delays, even in the presence of input delays, noise, and non-stationarity.
Compared to existing CPD methods, ODMD-CPD demonstrates competitive detection accuracy, robustness, and computational efficiency across synthetic benchmarks. Its deployment in real-world industrial-scale battery energy storage system, establishes ODMD-CPD as a practical solution for interpretable, real-time change-point detection in complex control environments.
