\section{Change-Point Detection Framework}\label{sec:method}

As discussed in Section~\ref{sec:preliminaries}, the success of identifying a low-rank subspace over which the signal evolves, while removing noise terms, relies on selecting the appropriate rank for the subspace. Projecting data onto modes of this low-rank subspace can result in increased reconstruction error when non-conforming patterns appear in the data. Transient dynamics, in particular, cannot be adequately captured by the low-rank subspace~\cite{Kuehn2011, Gottwald2020}. Therefore, a valid selection of the subspace maximizes the reconstruction error for non-stationary signals and is crucial for its use in CPD~\cite{Moskvina2003}.

Long-term deployment in systems with time-varying characteristics connected to factors such as aging, wear, or environmental conditions necessitates sequential detection and updates to the subspace in a streaming manner. This allows the system to adapt to slow changes in the time series structure and to accommodate new operations that may persist for an undefined duration. The ODMD-CPD algorithm is designed to address these challenges, providing a robust and adaptive solution for CPD in time series data.

Firstly, the algorithm projects two stored windows of snapshot pairs, referred to as base matrix \( \ui{\bar{X}}{B} \) and test matrix \( \ui{\bar{X}}{T} \), onto the subspace to evaluate the reconstruction error. Secondly, by comparing the reconstruction errors between the base and test matrices, the algorithm computes the change-point statistics. Finally, CPD-DMD updates the low-rank subspace over which the signal evolves using the new snapshots in base matrix.

\subsection{Mathematical Framework for Change-Point Detection}\label{sec:cpd-framework}

A change-point occurs when incoming data cannot be adequately reconstructed by the current subspace \( \tilde{\bar{A}}_k \), indicating a shift in the underlying system dynamics. Formally, we define the reconstruction error for an arbitrary augumented data matrix \(\bar{X}_k\) as:
\begin{equation}
\label{eq:reconstruction-error}
E(\bar{X}_k, \Phi_k) = \|\bar{X}_k - \Phi_k \Phi_k^T \bar{X}_k\|_F^2.
\end{equation}

Comparing reconstruction errors between a reference (base) window \(\ui{\bar{X}}{B}\) and a test window \(\ui{\bar{X}}{T}\), we compute the detection statistic \(Q_k\) as:
\begin{equation}
\label{eq:cpd-statistic}
Q_k = \max\left(0, \frac{E(\ui{\bar{X}}{T}, \Phi_k)/|\ui{\bar{X}}{T}|}{E(\ui{\bar{X}}{B}, \Phi_k)/|\ui{\bar{X}}{B}|} - 1\right)
\end{equation}
where \(|.|\) denotes the number of snapshots in given matrix.

Under the null hypothesis (no change), both windows should have similar reconstruction errors, yielding \(Q_k \approx 0\). Under the alternative hypothesis (change present), the test window will have significantly higher reconstruction error, resulting in \(Q_k > 0\). The magnitude of \(Q_k\) provides an interpretable measure of the change severity.

% \subsection{System Parameters}
\subsection{Data Stream Management}
Efficient execution of the algorithm requires preprocessing incoming data streams and managing the history of snapshots to compute the change-point statistics. Algorithm~\ref{alg:preprocessing} shows a single pass of the data preprocessing and management procedure, described below. This procedure is executed for each available snapshot pair or in mini-batches of varying frequency and size. First, incoming snapshots are formed into time-delayed embeddings of a predefined number of delays \( h \), as shown in Eq.~\eqref{eq:hankel}. Next, the time-delayed embedding of one-step delayed snapshots \( X^\prime_{h, k: k + j} \) is compensated by control action if the control matrix \( B \) is known, or the time-delayed embedding \( X_{h, k: k + j} \) is augmented with control actions \( \Theta_{h, k: k + j} \) to form the augmented matrix \( \bar{X}_{h, k: k + j} \).

Four parameters \(a, b, c, d\) define three required snapshot sets; the base set \({\ui{\bar{X}}{B} = \{x_h(t_i)\}}^{k - b - c}_{i=k - a - b - c}\), the test set \(\ui{\bar{X}}{T} = {\{x_h(t_i)\}}^{k}_{i=k - c}\), and the learning pair \( \{\ui{\bar{X}}{L}, \ui{\bar{X}^\prime}{L}\} = {\{x_h(t_i), x_h(t_i^\prime )\}}^{k - b - c + j}_{i=k - b - c - d}\). Conveniently, storing snapshots pairs \( \{\ui{\bar{X}}{all}, \ui{\bar{X}^\prime}{all}\} = {\{x_h(t_i), x_h(t_i^\prime )\}}^{k}_{i=k - b - c - d}\) is sufficient to manage all required data efficiently. In Section~\ref{sec:systematic-parameters}, we will explain the selection of these parameters.

% TODO: notation of the number of snapshots changed from c to j
\begin{algorithm}
	\caption{Single pass of data preprocessing and management procedure}\label{alg:preprocessing}
	\begin{algorithmic}[1]
		\REQUIRE{
			\(X_{k: k + j}\), % Incoming snapshots
			\(X^\prime_{k: k + j}\), % One-step delayed snapshots
			\(\Theta_{k: k + j}\), % Control actions
			\(\ui{\bar{X}}{all}\), % Stored snapshots
			\(\ui{\bar{X}^\prime}{all}\), % Stored delayed snapshots
			\(h\), % Number of delays
			\(B\), % Control matrix
			\(a, b, c, d\) % Parameters defining snapshot sets
		}
		\ENSURE{
			\(\ui{\bar{X}}{L}\), % Learning snapshots
			\(\ui{\bar{X}^\prime}{L}\), % Learning delayed snapshots
			\(\ui{\bar{X}}{B}\), % Base snapshots
			\(\ui{\bar{X}}{T}\), % Test snapshots
			\(\ui{\bar{X}}{all}\), % Updated stored snapshots
			\(\ui{\bar{X}^\prime}{all}\), % Updated stored delayed snapshots
			\(j\) % Number of snapshots
		}

		\STATE{\(j \gets \text{number of snapshots in } X_{k: k + j}\)}

		\STATE{\(X_{h, k: k + j} \gets \text{hankelize}(X_{k: k + j}, h)\)}
		\STATE{\(X^\prime_{h, k: k + j} \gets \text{hankelize}(X^\prime_{k: k + j}, h)\)}
		\STATE{\(\Theta_{h, k: k + j} \gets \text{hankelize}(\Theta_{k: k + j}, h)\)}

		\IF{\(B\) is known}
		\STATE{\(\bar{X}_{h, k: k + j} \gets X_{h, k: k + j}\)}
		\STATE{\(\bar{X}^\prime_{h, k: k + j} \gets X^\prime_{h, k: k + j} - B \Theta_{h, k: k + j}\)}
		\COMMENT{Eq.~\eqref{eq:control-compensation}}
		\ELSE{}
		\STATE{\(\bar{X}_{h, k: k + j} \gets \begin{bmatrix} X_{h, k: k + j} \\ \Theta_{h, k: k + j} \end{bmatrix}\)}
		\COMMENT{Eq.~\eqref{eq:augmented-matrix}}
		\STATE{\(\bar{X}^\prime_{h, k: k + j} \gets X^\prime_{h, k: k + j}\)}
		\ENDIF{}

		\STATE{\(\ui{\bar{X}}{L} \gets \ui{\bar{X}}{all}(k - b - c - d: k - b - c + j)\)}
		\STATE{\(\ui{\bar{X}^\prime}{L} \gets \ui{\bar{X}^\prime}{all}(k - b - c - d: k - b - c + j)\)}

		\STATE{\(\ui{\bar{X}}{all} \gets \begin{bmatrix} \ui{\bar{X}}{all}(j: k) & \bar{X}_{h, k: k + j} \end{bmatrix}\)}
		\STATE{\(\ui{\bar{X}^\prime}{all} \gets \begin{bmatrix} \ui{\bar{X}^\prime}{all}(j: k) & \bar{X}^\prime_{h, k: k + j} \end{bmatrix}\)}

		\STATE{\(\ui{\bar{X}}{B} \gets \ui{\bar{X}}{all}(k - a - b - c: k - b - c)\)}
		\STATE{\(\ui{\bar{X}}{T} \gets \ui{\bar{X}}{all}(k - c: k)\)}
	\end{algorithmic}
\end{algorithm}


\subsection{Learning Procedure}\label{learn-cpd}
The learning procedure involves updating the Dynamic Mode Decomposition (DMD) model with new snapshots and forgetting old ones to track the system's time-varying characteristics. This procedure is outlined in Algorithm~\ref{alg:learning}.

\begin{algorithm}
	\caption{Single pass of learning procedure of CPD-DMD}\label{alg:learning}
	\begin{algorithmic}[1]
		\REQUIRE{
			\(\ui{\bar{X}}{L}\),
			\(\ui{\bar{X}^\prime}{L}\),
			\(c\),
			\(j\)
		}
		% \ENSURE{\(\Phi \)}
		\STATE{\(\tilde{c} \gets \text{number of snapshots in } \ui{\bar{X}}{L} \)}
		\STATE{\(j^\prime \gets \tilde{c} + j - c \)}
		\COMMENT{\(\tilde{c} \leq c\) smaller before fully loaded}
		\IF{\(j^\prime > 0\)}
		\STATE{\textbf{Revert:} DMD\( (
			-\ui{\bar{X}}{L}(: j^\prime),
			-\ui{\bar{X}^\prime}{L}(: j^\prime)
			) \)}
		\COMMENT{Eq.~\eqref{eq:precision-matrix-update},~\eqref{eq:online-dmd-update}}
		\ENDIF{}
		\STATE{\textbf{Update:} DMD\( (
			\ui{\bar{X}}{L}(- j: ),
			\ui{\bar{X}^\prime}{L}(- j: )
			) \)}
		\COMMENT{Eq.~\eqref{eq:precision-matrix-update},~\eqref{eq:online-dmd-update}}
		% \STATE{\(\Phi \leftarrow \text{DMD}()\) \COMMENT{retrieve full modes}}
	\end{algorithmic}
\end{algorithm}

Initially, we verify the number of snapshots in the learning set and revert the DMD subspace if the learning set is fully loaded. Note that the learning set might not be full at the start of the learning procedure but must contain at least \((m + l) h\) snapshots, assuming unique measurements and that the learning set \(\ui{\bar{X}}{L}\) has full column rank. Subsequently, we update the DMD subspace with new snapshots entering the learning set. This procedure is repeated for each snapshot pair available or in mini-batches, whose frequency and size may not be uniform, governed by the upstream message queuing service.

\subsection{Detection Procedure}\label{detect-cpd}
The detection procedure, executed before the learning procedure, computes the change-point statistics. This sequence is crucial to avoid false negatives, as new snapshots may represent transient dynamics, and updating the DMD subspace beforehand could result in misidentification. Although the impact of this sequence is minimal for rank-one updates since both procedures are executed in the same pass, its importance grows with the relative size of the mini-batch to the potential span of the change-point. Nevertheless, aligning with best practices prevents any information leaks.

Algorithm~\ref{alg:detection} details the detection procedure. First, we project the base and test matrices onto the DMD subspace. Second, we reconstruct the full state representation and calculate the sum of squared Euclidean distances between the data and their DMD reconstruction. Third, we normalize this sum by the number of snapshots in the matrices. Finally, we compute the change-point statistics as the ratio of errors between the test and base matrices.

In cases where the error ratio is less than 1, the reconstructed test set \(\ui{\tilde{\bar{X}}}{T}\) captures more information about \(\ui{\bar{X}}{T}\) than the reconstructed base set \(\ui{\tilde{\bar{X}}}{B}\) about \(\ui{\bar{X}}{B}\). This rare scenario typically occurs when the signal is stationary, but the noise variance decreases in the test set compared to the training set. Although this phenomenon is interesting as it indicates a change in noise variance and the end of transient regime states, it is not considered in this paper, and we truncate the value to 1 and shift the score to zero, defining the minimum energy of matching errors.

\begin{algorithm}
	\caption{Single pass of detection procedure of CPD-DMD}\label{alg:detection}
	\begin{algorithmic}[1]
		\REQUIRE{
			\(\ui{\bar{X}}{B}\),  % Base matrix
			\(\ui{\bar{X}}{T}\),  % Test matrix
			\( \Phi \)            % DMD modes matrix
		}
		\ENSURE{
			\(Q_k\)  % Change-point statistics
		}
		\newline{\textbf{Project matrices onto DMD subspace}}
		\STATE{\(
			\ui{\tilde{\bar{X}}}{B} \gets \Phi^T \ui{\bar{X}}{B} \newline
			\ui{\tilde{\bar{X}}}{T} \gets \Phi^T \ui{\bar{X}}{T}
			\)}
		\newline{\textbf{Compute reconstruction errors}}
		\STATE{\(
			\ui{E}{B} \gets \sum_{i = k - a - b - c}^{k - b - c} \left \| \uis{\bar{X}}{B}{i} - \Phi \uis{\tilde{\bar{X}}}{B}{i} \right \|_F^2 \newline
			\ui{E}{T} \gets \sum_{i = k - c}^{k} \left \| \uis{\bar{X}}{T}{i} - \Phi \uis{\tilde{\bar{X}}}{T}{i} \right \|_F^2
			\)}
		\newline{\textbf{Normalize errors by the number of snapshots}}
		\STATE{\(
			\ui{E}{B} \gets \frac{\ui{E}{B}}{a} \newline
			\ui{E}{T} \gets \frac{\ui{E}{T}}{c}
			\)}
		\newline{\textbf{Compute change-point statistics}}
		\STATE{
			\(Q_k \gets \text{max}(0, \frac{\ui{E}{T}}{\ui{E}{B}} - 1)\)
		}
	\end{algorithmic}
\end{algorithm}


\subsection{Full Algorithm}
The complete CPD algorithm comprises three fundamental steps, as outlined in Algorithm~\ref{alg:cpd-dmd}. While the internal parameters of each step are abstracted for readability, their updates remain important. The proposed architecture is tailored for real-time execution, making it ideal for deployment in industrial environments characterized by dynamic data acquisition and irregular sampling patterns, often orchestrated by message queuing services.

\begin{algorithm}
	\caption{Single pass of CPD-DMD procedure}\label{alg:cpd-dmd}
	\begin{algorithmic}[1]
		\REQUIRE{
			\(X_{k: k + j}\),  % Current snapshots
			\(X^\prime_{k: k + j}\),  % One-step delayed snapshots
			\(\Theta_{k: k + j}\)  % Control actions
		}
		\ENSURE{
			\(Q_k\)  % Change-point statistics
		}
		\newline{\textbf{Step 1: Preprocessing the data}}
		\STATE{\(
			\ui{\bar{X}}{L}, \ui{\bar{X}^\prime}{L}, \ui{\bar{X}}{B}, \ui{\bar{X}}{T} \gets
			\text{preprocessing}( X_{k: k + j}, X^\prime_{k: k + j}, \Theta_{k: k + j})
			\)}
		\COMMENT{Refer to Algorithm~\ref{alg:preprocessing}}
		\newline{\textbf{Step 2: Detecting change-points}}
		\STATE{\(
			Q_k \gets \text{detection}( \ui{\bar{X}}{B}, \ui{\bar{X}}{T} )
			\)}
		\COMMENT{Refer to Algorithm~\ref{alg:detection}}
		\newline{\textbf{Step 3: Updating the learning model}}
		\STATE{\(
			\text{learning}( \ui{\bar{X}}{L}, \ui{\bar{X}^\prime}{L} )
			\)}
		\COMMENT{Refer to Algorithm~\ref{alg:learning}}

	\end{algorithmic}
\end{algorithm}

% \input{sections/method-parameter_selection}
